### 2014-06-27 manuscript/manuscript.tex
Final edits II.
Final edits III.

@@

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic study make illuminating this ``dark matter'' all the more
urgent. Among the remarkable discoveries made are: riboswitch-mediated
regulation \cite{Barrick:2007,Breaker:2012}, transcriptional
termination by RNA elements
\cite{von_Hippel:1998,Gardner:2011,Santangelo:2011}, identification of
novel natural catalytic RNAs
\cite{Kruger:1982,Guerrier-Takada:1983,Winkler:2004,Roth:2014},
CRISPR-mediated acquired immunity \cite{Barrangou:2007,Brouns:2008},
temperature-dependent gene regulation \cite{Narberhaus:2006,Loh:2013},
and sno-like RNAs in Archaea
\cite{Omer:2000,Gaspin:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset \cite{Wu:2009} did help improve
identification of broadly-conserved RNA families
\cite{Hoeppner:2012}. Tools such as RNA-seq \cite{Croucher:2010} and
transposon insertion sequencing
\cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a} promise to
complement comparative genomics tools for RNA family discovery, and it
may be possible to use a mix of data types in the identification of
RNA elements. However, to date, no systematic analysis of available
data has been undertaken, suggesting ncRNAs may be hidden in the
deluge of published data.

@@

We next calculated the percentage of conserved RUFs for all
Enterobacterial species pairs. On average, 83\% of RUFs are conserved
across the Goldilocks Zone. The two \emph{E. coli} strains are extremely
similar, and share 99\% of their RUFs, suggesting that these species
are too similar for us to robustly separate expression of \emph{bona fide}
RNAs and noise. While these outputs could be genuine RNAs, these
strains are in the ‘too hot’ region, meaning if everything is
conserved, comparative power is lost. In contrast, only 12\% of RUFs
are conserved between species pairs in the ‘too cold’ region (spanning
clades; Figure~\ref{fig:3}) and of the 197 RUFs found through
comparative analysis of transcriptomes within the Goldilocks Zone,
only 19 show evidence of expression in another transcriptome outside
of this zone. This suggests that the low number of RUFs from
Figure~\ref{fig:2}D showing both conservation and expression is
primarily a consequence of limited sampling.  We also extracted a RUF
in the Enterobacterial clade that had homologs in the other four
species.  That said, mining RNA-seq data within the Goldilocks Zone
permits a higher confidence in the identification of novel ncRNAs. One
such example of this is illustrated in (Figure~\ref{fig:4}B\&C). This
RUF exhibits sequence and secondary structure conservation and is
expressed at high levels across all five Enterobacterial
transcriptomes in our dataset.

@@

All available bacterial and archaeal genomes were downloaded from the
European Nucleotide Archive (ENA) (2,562 and 154 genomes,
respectively) \cite{Cochrane:2013}. RNA-seq datasets published as of
March 2013 were collected, spanning 37 species/strains, 44 experiments
and 413 lanes of sequencing data (Table S1). Most of these datasets
are generated on the Illumina platform \cite{Shendure:2005}, with a
few lanes from the SOLiD platform \cite{Cloonan:2008} and the 454
platform \cite{Margulies:2005}. Where possible, FastQ files were
downloaded, scanned for residual adapter sequences using
AdapterRemoval \cite{Lindgreen:2012}, and mapped to the reference genome
using Bowtie2 \cite{Langmead:2012} for Illumina and 454 data
and BFAST \cite{Homer:2009} for SOLiD data.

@@

Homologs of the identified RUFs were found in all the downloaded
genomes using nhmmer \cite{Wheeler:2013} in an iterative fashion:
First, the RUF sequence alone was used in the scan; then, all hits
with $E-value<0.001$ were included and a HMM built. This was iterated
5 times. The alignments from the RUF homology search were analyzed
further by investigating the potential for secondary structure
formation using RNAz \cite{Gruber:2010} and alifoldz
\cite{Washietl:2004}. Protein coding potential of the RUFs was
assessed using RNAcode \cite{Washietl:2011}. Overlaps between
potential RUF homologs in other species and all the annotations in the
respective genomes was also assessed.

@@

{\bf Expression of core genes:} We defined a set of 40 core
protein-coding genes based on \cite{Wu:2013,Darling:2014} and 16 noncoding RNA
genes (the union of tRNA, RNaseP, tmRNA, SRP, 6S and rRNA RNA
families) \cite{Gardner:2011,Burge:2013}. If the median read depth
greater than 10X, we defined the gene as expressed. For each dataset,
we report the fraction of the 40 core genes that are expressed.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure1.pdf}
\end{center}
\caption{ {\bf Identification of transcribed elements across
    publicly-available RNA-seq data.} Non-coding RNA elements show
  high expression across transcriptomes. Both annotated Rfam families
  (red – core Rfam families are dark red, all others are light red)
  and expressed RUFs (black) are among the highest expressed outputs
  in transcriptomes. Protein-coding regions annotated using Pfam are
  also shown (blue). For each species we generated relative rankings
  of expression spanning protein coding genes, RNA genes and candidate
  RUF genes. Genes were extracted from mapping data by comparing to
  the annotation. Going from mapped reads to expression level can be
  confounded by a number of factors (e.g. sample preparation, overall
  sequencing depths, rRNA depletion etc.). For consistency, we
  therefore ranked genes for each species and compared rankings
  instead of comparing the read depths directly between species. For a
  given species, the annotated genes were ranked based on the median
  read depth of the annotated region. RUFs were manually picked by
  masking out annotated genes and selecting regions showing evidence
  of expression by inspecting read depth across the genome. This
  yielded 844 gene candidate sequences in Bacteria (78 in
  Archaea). The plot contains the 50 most highly expressed elements
  for each species.}
\label{fig:1}
\end{figure}
%pdftops -eps figures/figure1.pdf figures/figure1.eps

>>>>

%Anthony Poole <anthony.poole@canterbury.ac.nz>, Anthony Poole <poulet.m@gmail.com>, Hisham ElDai <biohisham@gmail.com>, Hisham ElDai <hisham.eldai@pg.canterbury.ac.nz>, Lars Barquist <lb14@sanger.ac.uk>, Lars Barquist <lbarquist@gmail.com>, Sinan Ugur Umu <sinan.umu@pg.canterbury.ac.nz>, Sinan Ugur Umu <sinanugur@gmail.com>, Sook-Wei Lai Alicia <alicialsw85@gmail.com>, Stinus <stinus.lindgreen@canterbury.ac.nz>, Stinus <stinus.lindgreen@gmail.com>, stinus@binf.ku.dk,

@@

The paucity of annotations is understandable, as RNA gene annotation
is non-trivial \cite{Freyhult:2007,Nawrocki:2009}. However, the
increasing roles for RNA uncovered through experimental and
bioinformatic studies make illuminating this ``dark matter'' all the more
urgent. Among the remarkable discoveries made are: riboswitch-mediated
regulation \cite{Barrick:2007,Breaker:2012}, transcriptional
termination by RNA elements
\cite{von_Hippel:1998,Gardner:2011,Santangelo:2011}, identification of
novel natural catalytic RNAs
\cite{Kruger:1982,Guerrier-Takada:1983,Winkler:2004,Roth:2014},
CRISPR-mediated acquired immunity \cite{Barrangou:2007,Brouns:2008},
temperature-dependent gene regulation \cite{Narberhaus:2006,Loh:2013},
and sno-like RNAs in Archaea
\cite{Omer:2000,Gaspin:2000,Gardner:2010}. The Rfam database
\cite{Gardner:2011,Burge:2013} provides a valuable platform for
collating and characterising these and other families of noncoding
RNA. However, a recent comparative analysis \cite{Hoeppner:2012}
revealed that fewer than 7\% of RNA families within Bacteria and less
than 19\% in Archaea show a broad phylogenetic distribution (that is,
presence in at least 50\% of sequenced phyla). Crucially, that
analysis revealed that underlying genome sequencing biases were a
major contributor to this pattern, and that the wider genomic sampling
provided by the GEBA dataset \cite{Wu:2009} did help improve
identification of broadly-conserved RNA families
\cite{Hoeppner:2012}. Tools such as RNA-seq \cite{Croucher:2010} and
transposon insertion sequencing
\cite{van_Opijnen:2013,Barquist:2013,Barquist:2013a} promise to
complement comparative genomics tools for RNA family discovery, and it
may be possible to use a mix of data types in the identification of
RNA elements. However, to date, no systematic analysis of available
data has been undertaken, suggesting ncRNAs may be hidden in the
deluge of published data.

@@

We next calculated the percentage of conserved RUFs for all
Enterobacterial species pairs. On average, 83\% of RUFs are conserved
across the Goldilocks Zone. The two \emph{E. coli} strains are extremely
similar, and share 99\% of their RUFs, suggesting that these species
are too similar for us to robustly separate expression of \emph{bona fide}
RNAs and noise. While these outputs could be genuine RNAs, these
strains are in the ‘too hot’ region, meaning if everything is
conserved, comparative power is lost. In contrast, only 12\% of RUFs
are conserved between species pairs in the ‘too cold’ region (spanning
clades; Figure~\ref{fig:4}) and of the 197 RUFs found through
comparative analysis of transcriptomes within the Goldilocks Zone,
only 19 show evidence of expression in another transcriptome outside
of this zone. This suggests that the low number of RUFs from
Figure~\ref{fig:2}D showing both conservation and expression is
primarily a consequence of limited sampling.  We also extracted a RUF
in the Enterobacterial clade that had homologs in the other four
species.  That said, mining RNA-seq data within the Goldilocks Zone
permits a higher confidence in the identification of novel ncRNAs. One
such example of this is illustrated in (Figure~\ref{fig:4}B\&C). This
RUF exhibits sequence and secondary structure conservation and is
expressed at high levels across all five Enterobacterial
transcriptomes in our dataset.

@@

All available bacterial and archaeal genomes were downloaded from the
European Nucleotide Archive (ENA) (2,562 and 154 genomes,
respectively) \cite{Cochrane:2013}. RNA-seq datasets published as of
August 2013 were collected, spanning 37 species/strains, 44 experiments
and 413 lanes of sequencing data (Table S1). Most of these datasets
are generated on the Illumina platform \cite{Shendure:2005}, with a
few lanes from the SOLiD platform \cite{Cloonan:2008} and the 454
platform \cite{Margulies:2005}. Where possible, FastQ files were
downloaded, scanned for residual adapter sequences using
AdapterRemoval \cite{Lindgreen:2012}, and mapped to the reference genome
using Bowtie2 \cite{Langmead:2012} for Illumina and 454 data
and BFAST \cite{Homer:2009} for SOLiD data.

@@

Homologs of the identified RUFs were found in all the downloaded
genomes using nhmmer \cite{Wheeler:2013} in an iterative fashion:
First, the RUF sequence alone was used in the scan; then, all hits
with E-value$<0.001$ were included and a HMM built. This was iterated
5 times. The alignments from the RUF homology search were analyzed
further by investigating the potential for secondary structure
formation using RNAz \cite{Gruber:2010} and alifoldz
\cite{Washietl:2004}. Protein coding potential of the RUFs was
assessed using RNAcode \cite{Washietl:2011}. Overlaps between
potential RUF homologs in other species and all the annotations in the
respective genomes was also assessed.

@@

{\bf Expression of core genes:} We defined a set of 40 core
protein-coding genes based on \cite{Wu:2013,Darling:2014} and 16 noncoding RNA
genes (the union of tRNA, RNaseP, tmRNA, SRP, 6S and rRNA RNA
families) \cite{Gardner:2011,Burge:2013}. If the median read depth is
greater than 10X, we defined the gene as expressed. For each dataset,
we report the fraction of the 40 core genes that are expressed.

@@

\begin{figure}[!ht]
\begin{center}
\includegraphics[width=4in]{figures/figure1.pdf}
\end{center}
\caption{ {\bf Identification of transcribed elements across
    publicly-available RNA-seq data.} Non-coding RNA elements show
  high expression across transcriptomes. Both annotated Rfam families
  (red - core Rfam families are dark red, all others are light red)
  and expressed RUFs (black) are among the highest expressed outputs
  in transcriptomes (blue - core Pfam families are dark blue, all
  others are light blue). For each species we generated relative
  rankings of expression spanning protein coding genes, RNA genes and
  candidate RUF genes. Genes were extracted from mapping data by
  comparing to the annotation. Going from mapped reads to expression
  level can be confounded by a number of factors (e.g. sample
  preparation, overall sequencing depths, rRNA depletion etc.). For
  consistency, we therefore ranked genes for each species and compared
  rankings instead of comparing the read depths directly between
  species. For a given species, the annotated genes were ranked based
  on the median read depth of the annotated region. RUFs were manually
  picked by masking out annotated genes and selecting regions showing
  evidence of expression by inspecting read depth across the
  genome. This yielded 844 gene candidate sequences in Bacteria (78 in
  Archaea). The plot contains the 50 most highly expressed elements
  for each species.}
\label{fig:1}
\end{figure}
%pdftops -eps figures/figure1.pdf figures/figure1.eps
